---
title: "Tables"
format: html
editor: visual
---

# Tables

This document will go over the specifications for converting R data frames
and their associated operations into PMML.

## Context

Certain algorithms requires the use of tables to store reference information
that is needed during the calculation. For example, the sodium plus algorithm
has a reference table of sodium per serving for different food groups stored
in a CSV file; the diet calculator has multiple lookup tables that contain
the different equation values to use for the different food groups.

These tables are imported into the R code, filtered on different conditions,
and finally a value selected from a column to use in the calculation. All of
this information needs to be included in the PMML file so that the algorithm
can be full represented in it as well as allowing any PMML engine to score
the algorithm.

## R Tables

This section will go over the different pieces of R code that involve table
operations and the PMML XML they should be converted to.

### Importing a table

R code can be used to import a CSV file into a table. In R, a table is
represented using a structure called data frame. In R, data frames can either
be created by using the data.frame constructor or by importing a tabular file
like a CSV into a data frame. Regardless of the way a data frame is created,
the PMML nodes, `Taxonomy`, `InlineTable`, and `row` are used to represent
the data within the data frame.

The `Taxonomy` node defines the name of the table. The `name` attribute should
be set to the name of the variable.

The `InlineTable` node contains the table contents within it. The only children
allowed are `row` nodes.

The `row` nodes are used to represent each row in the table. Each child node
represents a column in that row. The child node name has the column name and
the contents has the column value.

For example, for the following data frame code,

```{r}
table <- data.frame(
  a = c(1,2,3),
  b = c(4,5,6)
)
```

the following PMML string should be generated,

```{xml}
<Taxonomy name="table">
  <InlineTable>
    <row>
      <a>1</a>
      <b>4</b>
    </row>
    <row>
      <a>2</a>
      <b>5</b>
    </row>
    <row>
      <a>3</a>
      <b>6</b>
    </row>
  </InlineTable>
</Taxonomy>
```

One issue is deciding when to convert a table variable to a `Taxonomy` node
and when to use a `DerivedField` node.

Usually, a `Taxonomy` node is used when tables are created for the first time.
Currently, we would only support the following table creation expression:

1. A table created by using the `read.csv` function

In all the other cases, where a table variable is used to create a new table
variable, we would create a `DerivedField` element as shown in the next section.

## Filtering & Selecting

R allows the user to filter the rows in a table based on a set of conditions
as well as select the values from a column(s).

### Filtering

Example R code to filter rows from a table can be seen below,

```{r}
filtered_table <- table[table$a == 1, ]
```

The condition within the above R code says to only keep rows whose `a` column
has a value of `1`. 

Although this is a simple condition, R allows you to define complex conditions, 
essentially any statement that evaluates to a boolean can be used as the filter 
logic. However, due to limitations with PMML, the conversion will only allow
conditions of the form, **equality_condition_1 & equality_condition_2 & .. & equality_condition_n**.
In English, only equality conditions are allowed, and if there are multiple
equality conditions, they have to all be seperated by an **AND**.

For example, the following filter condition is allowed,

```{r}
b <- 4
filtered_table_1 <- table[table$a == 1 & table$b == b, ]
```

There are two equality conditions and they are both seperated by an AND.

The following condition is not allowed,

```{r}
filtered_table_2 <- table[table$a != 1 | table$b != 4, ]
```

The two sub conditions not equalities (they are in-equalities) and they are
not seperated by an AND (they are seperated by an OR).

Converting an R filter statement to PMML uses the following elements, 

* `MapValues`: whose children contains the equality conditions
* `FieldColumnPair`: which encapsulates the equality condition. The `column`
attribute contains the name of column part of the filter, in other words the
left-hand side of the equality. 
Depending on what the right-hand side of the
equality is, either the `constant` or `field` attribute will need to be set.
If the right-hand side is a constant then the former is used and if it is a
variable the latter is used.

For example. for the two valid filter statements above, the equivalent
PMML is shown below,

```{xml}
<DerivedField name="filtered_table">
  <MapValues>
    <FieldColumnPair column="a" constant="1" />
  <MapValues>
</DerivedField>
```

```{xml}
<DerivedField name="b">
  <Constant>4</Constant>
</DerivedField>
<DerivedField name="filtered_table_1">
  <MapValues>
    <FieldColumnPair column="a" constant="1" />
    <FieldColumnPair column="b" field="b" />
  <MapValues>
</DerivedField>
```

### Selecting

R code can also be used to select values from a table's columns and use them
in another expression. For example, the code below will select the values from
the `a` column,

```{r}
a <- filtered_table$a
```

In addition, filtering and selecting can be combined in a single expression,

```{r}
a <- table[table$a == 1, ]$a
```

R also allows users to select multiple columns from a database, however PMML
does not support, meaning that neither does the software. The following
expression should return an error,

```{r}
ab <- table[table$a == 1, c("a", "b")]
```

Writing the equivalent PMML code is the same as for filtering, except we set
the `outputColumn` attribute to the selected column. The equivalent PMML
for the above two code chunks are shown below,

```{xml}
<DerivedField name="a">
  <MapValues outputColumn="a">
  </MapValues>
</DerivedField>
```

```{xml}
<DerivedField name="a">
  <MapValues outputColumn="a">
    <FieldColumnPair column="a" constant="1" />
  </MapValues>
</DerivedField>
```

Notice how in the first chunk there aren't any `FieldColumnPair` elements
since there aren't any conditions.

### `TableLocator`

In all the PMML chunks above we never specified where the actual table we're
filtering and selecting is coming from. This is where the `TableLocator` node
comes in, its used to specify where the table is coming from as well as the name
of the table to use from that location. If has the following attributes:

* `name`: The name of the table to use
* `location`: Where to look for the table. This can have the following values:
  * `taxonomy`: Look for it in the `Taxonomy` element within the PMML file
  * `local`: Look for it in a local variable, either in the `DerivedField` 
  element or `DataDictionary` element
  
The main issue is whether to set `location` to `taxonomy` or `local`. When 
deciding this value:

1. We first check if the table is present as a `Taxonomy`
element
2. If it isn't then we check if it is in the local scope, setting it to `local`
if it is
3. If it isn't local, then we return an error

For example, for the code chunks above that use the `table` variable, we would
use `taxonomy` as shown below,

```{xml}
<DerivedField name="filtered_table">
  <MapValues>
    <FieldColumnPair column="a" constant="1" />
    <TableLocator name="table" location="taxonomy" />
  <MapValues>
</DerivedField>
```

```{xml}
<DerivedField name="b">
  <Constant>4</Constant>
</DerivedField>
<DerivedField name="filtered_table_1">
  <MapValues>
    <FieldColumnPair column="a" constant="1" />
    <FieldColumnPair column="b" field="b" />
    <TableLocator name="table" location="taxonomy" />
  <MapValues>
</DerivedField>
```

```{xml}
<DerivedField name="a">
  <MapValues outputColumn="a">
    <FieldColumnPair column="a" constant="1" />
    <TableLocator name="table" location="taxonomy" />
  </MapValues>
</DerivedField>
```

and for the remaining tables we would use `local` as shown below,

```{xml}
<DerivedField name="b">
  <Constant>4</Constant>
</DerivedField>
<DerivedField name="filtered_table_1">
  <MapValues>
    <FieldColumnPair column="a" constant="1" />
    <FieldColumnPair column="b" field="b" />
    <TableLocator name="filtered_table" location="local" />
  <MapValues>
</DerivedField>
```

#### `DefineFunction`

When working with a function, the situation is reversed:

1. We first check if the table is present as a local variable within the 
function's scope, setting the location to `local` if so
2. Otherwise we check if the table is present in a `Taxonomy` element, setting
it to `taxonomy` if so
3. Otherwise, we return an error

For example, look at the following two code chunks,

```{r}
table_func_1 <- function(table) {
  return(table$a)
}
```

```{r}
table_func_2 <- function() {
  return(table$a)
}
```

The first one has `table` variable that is passed in as an argument whereas
the second one is using the earlier defined `table`. The PMML for the two
code chunks are shown below,

```{xml}
<DefineFunction name="table_func_1">
    <ParameterField name="table" />
    <MapValues outputColumn="a">
      <TableLocator name="table" location="local" />
    </MapValues>
</DefineFunction>
```


```{xml}
<DefineFunction name="table_func_2">
    <MapValues outputColumn="a">
      <TableLocator name="table" location="taxonomy" />
    </MapValues>
</DefineFunction>
```
